Fail Faster

### 2025-03-19 paper/paper.tex
progress

\subsubsection{Uses of Staging in eDSL Construction}
One of the primary uses of staging is in embedded DSL construction \cite{sheard99, lms, trattdsl}.
While embedded DSLs are great \jwc{words}, they have well-known drawback: 
the functional abstractions used to build eDSLs prevent compilers from generating efficient machine code \cn.
As alluded to previously \jwc{This sentence here because of the section-ordering transpose... may need to mention this earlier.}, this effect is known as ``abstraction overhead'' \cite{carrette05, moller20}: the layers of abstraction that 
make eDSLs so usable are precisely what prevents them from being fast.
\jwc{Something to the effect of ``the issues we noticed in BQ and SC exist across many eDSLs''}
In light of this issue, staging is often used as a lightweight compiler for DSLs. The compile-time evaluation stage
transforms the DSL code, eliminating abstractions to produce code that the host language compiler can generate fast machine code for.
\jwc{This previous paragraph is peak ``shitty first draft''.}
This recipe has been used to great effect across domains to stage eDSLs for stream processing \cite{moller20,strymonas}, parser combinators \cite{sspc, staged-parsers, krishnaswami19, flap},
and query processing \cite{rhyme}.

Many languages have some degree of staging functionality, though support varies
from full native support (Scala 3 \cite{scalamacros}, Haskell \cite{templatehaskell}) to compiler extensions (OCaml \cite{metaocaml,macocaml}, Java \cite{mint}).
Moreover, many more languages have lighter-weight staging-like metaprogramming support in the form of macros.
See Section~\ref{subsection:other-langs} for a full discussion of just how portable these ideas are. \jwc{this sentence is a placeholder lol}

@@

\subsection{Staged Type-Derived Generators}

\jwc{
  \begin{itemize}
    \item In PBT in practice, we learned that in many cases, programmers don't
    even write custom generators, instead relying on type-derived generators.
    \item These generators just produce arbitrary values of a given type, not necessarily enforcing validity conditions.
    \item If 
    \item Let's talk about how type-deriving works. In languages with typeclasses, it works
    by typeclass resolution. In OCaml it works by PPX system, but the principle
    is the same. You define rules to go from generators of a subtypes to a generator of the larger type, and then apply those rules to build up a full generator.
    \item \begin{itemize}
      \item For base types, you just call the associated generator
      \item For product types, you sample from all the component gnerators, bind the values, and then tuple up the values, and return the tuple
      \item For variant types, you use a weighted union to choose one of the component generators with a weighted union.
      \item For recursive types, wrap the whole thing in a fixedpoint and then use the recusive handle as the ``component generator'' for all recursive instance of the type.
    \end{itemize}
    \item This kind of generic deriving of generators works just as well for
    staged generators. Just replace the standard combinators in question with
    staged ones!
    \item We built this in OCaml, but you can easily use typeclasses to do it in scala too.
    \item Note that this is (essentially) 3-stage metaprograming. You're using either the PPX mechanism or typeclass resolution to generate a bunch of staged combinator calls, which then run at compile time, which then run at run time.
  \end{itemize}
}
\jwc{Todo: Thia}

@@

\subsection{Benchmarking Generator Speed}

\begin{figure}[h]
  % \includegraphics[scale=0.33]{allocs.png}
  \includegraphics[scale=0.45]{time.png}
  \label{fig:time}
  \caption{Average generation time over 10,000 trials. Across all benchmarks, we see reliable improvement from at least one of Staged SR and Staged CSR.}
\end{figure}

@@

It is worth noting that the time-to-failure of a given strategy on a given seed is not necessarily representative of 
the strategy's average time-to-failure over a large number of trials. We normalize by computing the relative performance of \name generators vs. \bq generators (the ``speedup''). While this works in most cases,
it is not perfect.
For example, it is theoretically possible
to choose a seed such that the \textit{first} value produced by a generator discovers the bug, which would elide any differential speedup (and likewise
if both generators fail to find the bug).
To account for these edge cases, we repeat the above process for 10 seeds and compute the geometric average of the speedups.\footnote{Although the variance \textit{between} seeds in Etna can be vast, timing results are replicable for a \textit{given} seed. Across 1,000 trials using the same seed, 
the observed variance in time-to-failure was less than a nanosecond.}

@@

In the years since the original QuickCheck paper \cn{}, PBT has had remarkable success
in languages outside of

Of course, users and developers of PBT libraries in \emph{other} languages might be curious about how
the techniques brought to bear on OCaml's \bq{} and Scala's ScalaCheck in this paper might be imported to their favorite language.

Indeed, while we have done our best to keep the proceeding discussion as
language-agnostic as possible, our claims about the \name{} technique have only
been thoroughly tested in OCaml and Scala 3, and we are only claiming (in the
scientific sense) that we have contributed a technique applicable to PBT
libraries in strict functional programming languages.

In this section

@@

\subsection{Staging Monadic Libraries}
\jwc{A generator is a parser of
randomness~\cite{goldsteinParsingRandomness2022}, and people have implemented
lots of staged parser libraries!}

\section{Conclusion \& Future Work}
\jwc{Why we stopped where we stopped: can you squeeze more out of this?}

>>>>

\subsubsection{Uses of Staging in eDSL Construction}
One of the primary uses of staging is in embedded DSL construction \cite{sheard99, lms, trattdsl}.
\jwc{Cite racket macros for DSLs here:}
While embedded DSLs are great \jwc{words}, they have well-known drawback: 
the functional abstractions used to build eDSLs prevent compilers from generating efficient machine code \cn.
\jwc{Do racket people also use macros to eliminate run time overhead? Or is it just for extensibility?}
As alluded to previously \jwc{This sentence here because of the section-ordering transpose... may need to mention this earlier.}, this effect is known as ``abstraction overhead'' \cite{carrette05, moller20}: the layers of abstraction that 
make eDSLs so usable are precisely what prevents them from being fast.
\jwc{Something to the effect of ``the issues we noticed in BQ and SC exist across many eDSLs''}
In light of this issue, staging is often used as a lightweight compiler for DSLs. The compile-time evaluation stage
transforms the DSL code, eliminating abstractions to produce code that the host language compiler can generate fast machine code for.
\jwc{This previous paragraph is peak ``shitty first draft''.}
This recipe has been used to great effect across domains to stage eDSLs for stream processing \cite{moller20,strymonas}, parser combinators \cite{sspc, staged-parsers, krishnaswami19, flap},
and query processing \cite{rhyme}.
\jwc{Cite racket things here}

Many languages have some degree of staging functionality, though support varies
from full native support (Scala 3 \cite{scalamacros}, Haskell \cite{templatehaskell}, Racket \cn{}) to compiler extensions (OCaml \cite{metaocaml,macocaml}, Java \cite{mint}).
Moreover, many more languages have lighter-weight staging-like metaprogramming support in the form of macros.
See Section~\ref{subsection:other-langs} for a full discussion of just how portable these ideas are. \jwc{this sentence is a placeholder lol}

@@

\subsection{Staged Type-Derived Generators}
\label{subsection:type-derived}
\jwc{
  \begin{itemize}
    \item In PBT in practice, we learned that in many cases, programmers don't
    even write custom generators, instead relying on type-derived generators.
    \item These generators just produce arbitrary values of a given type, not necessarily enforcing validity conditions.
    \item If 
    \item Let's talk about how type-deriving works. In languages with typeclasses, it works
    by typeclass resolution. In OCaml it works by PPX system, but the principle
    is the same. You define rules to go from generators of a subtypes to a generator of the larger type, and then apply those rules to build up a full generator.
    \item \begin{itemize}
      \item For base types, you just call the associated generator
      \item For product types, you sample from all the component gnerators, bind the values, and then tuple up the values, and return the tuple
      \item For variant types, you use a weighted union to choose one of the component generators with a weighted union.
      \item For recursive types, wrap the whole thing in a fixedpoint and then use the recusive handle as the ``component generator'' for all recursive instance of the type.
    \end{itemize}
    \item This kind of generic deriving of generators works just as well for
    staged generators. Just replace the standard combinators in question with
    staged ones!
    \item We built this in OCaml, but you can easily use typeclasses to do it in scala too.
    \item Note that this is (essentially) 3-stage metaprograming. You're using either the PPX mechanism or typeclass resolution to generate a bunch of staged combinator calls, which then run at compile time, which then run at run time.
  \end{itemize}
}
\jwc{Todo: Thia}

@@

\subsection{Benchmarking Generator Speed}
To answer \textbf{RQ1}, we implement benchmarks in both \name and \bq. We vary the choice of \rand in our \name generators, 
using both \bq's default \texttt{splittable\_random} (``SR'') and our improved SplitMix (``CSR''), as discussed in \secref{subsection:faster-rng}.

Our benchmarks consist of lists, binary search trees (BST), and simply typed lambda calculus (STLC) terms. We implement generators for these benchmarks using a variety of 
\textit{strategies}, which vary in structure and sophistication. These include: a derived generators for BSTs and STLC terms, using the approach outlined in \secref{subsection:type-derived}; 
two custom BST generators, one that repeatedly inserts values into an initially empty tree and one that materializes the tree in a single pass by generating keys, values, and subtrees at each step;
a custom, correct by construction generator for STLC, which is the most complex; and a generator of boolean lists.

\begin{figure}[h]
  % \includegraphics[scale=0.33]{allocs.png}
  \includegraphics[scale=0.45]{time.png}
  \label{fig:time}
  \caption{Average generation time over 10,000 trials.}
\end{figure}

@@

It is worth noting that the time-to-failure of a given strategy on a given seed is not necessarily representative of 
the strategy's average time-to-failure over a large number of trials. We normalize by computing the \textit{relative} performance (the ``speedup''). While this works in most cases,
it is not perfect.
For example, it is theoretically possible
to choose a seed such that the \textit{first} value produced by a generator discovers the bug, which would elide any differential speedup (and likewise
if both generators fail to find the bug).
To account for these edge cases, we repeat the above process for 10 seeds and compute the geometric average of the speedups.\footnote{Although the variance \textit{between} seeds in Etna can be vast, timing results are replicable for a \textit{given} seed. Across 1,000 trials using the same seed, 
the observed variance in time-to-failure was less than a nanosecond.}

@@

In the years since the original QuickCheck paper \cn{}, PBT has had remarkable success
in languages outside of the small ICFP world from which it came.
For this reason, readers who are users and developers of PBT libraries in \emph{other} languages ---
including non-strict or procedural ones --- might be curious about how the
techniques brought to bear on OCaml's \bq{} and Scala's ScalaCheck in this paper
might be imported to their favorite language.
Below we discuss how \name{} could be applied in other languages.

% While we have done our best to keep the proceeding discussion as
% language-agnostic as possible, our claims about the \name{} technique have only
% been thoroughly tested in OCaml and Scala 3, and we are only claiming (in the
% scientific sense) that we have contributed a technique applicable to PBT
% libraries in strict functional programming languages.

\begin{itemize}
  \item \textbf{Racket} is the next target that we intend to test the \name{} methodology on. The entire Racket philosophy is intertwined with
  using macros to build small eDSLs like the ones we use to write generators, and so it is a natural fit.
  We have not yet tested the technique in Racket for the simple reason that the authorship team of this paper does not posess
  much Racket expertise \jwc{can we say this without unblinding?}.

  \item \textbf{Haskell} birthed PBT, and so it is natural to ask why we have not yet applied \name{} to it. The basic answer is that
  because most tasks in Haskell are accomplished with monads, GHC is designed specially to eliminate the run-time overhead of monadic
  abstractions! Indeed, because Haskell is pure, GHC can aggressively inline and beta-reduce
  nearly anything as part of its normal optimiation steps.
  For complex enough programs however, the heuristics GHC uses can degrade, leading to performance overheads for monadic code.
  For this reason, Haskell does have a multi-stage programming system called Template Haskell \cn{}, which is sometimes 
  to build eDSLs in contexts where one does not want to rely on the optimizer's heuristics.
  \item \textbf{Rust:} asdf
  \item \textbf{Python:} asdf
\end{itemize}

@@

\subsection{Staging eDSLs}
\jwc{A generator is a parser of
randomness~\cite{goldsteinParsingRandomness2022}, and people have implemented
lots of staged parser libraries!}
\jwc{cite Lionel Parreaux  work, he's on PC}

\section{Conclusion \& Future Work}
\jwc{Why we stopped where we stopped: can you squeeze more out of this?}
\jwc{
  \begin{itemize}
    \item On the staging front, we could do a little with more clever staging to codegen better... Avoided using floating let-insertion to ensure we had precise control over effect ordering, but we could try to relax that.
          Recursive generators have some extra overhead for packing/unpacking accumulators, but could do a GADT cf \texttt{SOP} in 2ltt.
    \item On the RNG front, there's not much more we can squeeze out of the RNG without breakning equiv with splitmix. As discussed, there are much faster things you can do as alterative sources of entropy.
  \end{itemize}
}
